The function--sheaf dictionary for smooth commutative group schemes over finite fields with finitely generated geometric component group

### 2014-01-22 dictionary.tex
Fiddling with dictionary.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the sheaf--function dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question.

In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $\bG$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

@@

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to disconnected algebraic groups over finite fields. Consider the etale group scheme $G = \ZZ_{\Fq}$ over $\Fq$; this is a $\QC$-scheme.
Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. For example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is bounded but not finite, while the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$. This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

We also want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Some people refer to local systems satisfying \eqref{introbox} as as character sheaves; 
see for example, \cite{kamgarpour:09a}. 
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig. 
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

>>>>

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the sheaf--function dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question. 
%
In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $\bG$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

@@

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to $\QC$-schemes. Consider the etale group scheme $G = \ZZ_{/\Fq}$. Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. For example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is bounded but not finite, while the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, $\ell$-adic local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$. This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

We also want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Some people refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as as character sheaves; 
see for example, \cite{kamgarpour:09a}. 
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig. 
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.
